---
title: "Ancestral State Reconstruction"
output:
  pdf_document: default
  html_document: default
---

```{r setup, echo = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# General data analysis packages
library(knitr)
library(tidyverse)
library(magrittr)
library(plotrix)
library(plotly)

# Phylogenetics packages
library(ape)
library(phylobase)
library(phangorn)
library(phytools)
library(tidyverse)
library(ggtree)

```

based on (this tutorial)[http://www.phytools.org/eqg2015/asr.html].

# Get and clean files

```{r}

# read in tree and csv
fire.tree<-read.tree("fire.treefile")
fire.data<-read.csv("fire.csv", row.names = 1)
basics.tree<-read.tree("basics.treefile")
basics.data<-read.csv("basics.csv", row.names = 1)

# bookkeeping
## (checked for duplicate tips in python before creating .nex and .csv files!)

## make sure rows of character data are in same order as tip labels
fire.data = fire.data[order(match(row.names(fire.data), fire.tree$tip.label)), ]
basics.data = basics.data[order(match(row.names(basics.data), basics.tree$tip.label)), ]

## double check that labels are the same, in case some were missing or duplicated
if(! all( row.names(fire.data)==fire.tree$tip.label ) ){
  stop("fire.tree tip labels and dataframe rows not in same order!")
}
if(! all( row.names(basics.data)==basics.tree$tip.label ) ){
  stop("basics.tree tip labels and dataframe rows not in same order!")
}

## resolve polytomies
fire.tree <- multi2di(fire.tree)
basics.tree <- multi2di(basics.tree)

## root fire.tree by calculating midpoint
fire.tree <- midpoint(fire.tree)
## root basics.tree at mew
basics.tree <- root(basics.tree, "mew", resolve.root = TRUE)


# find and set important metadata abt each tree
## count number of tips
ntips_fire = length(fire.tree$tip.label)
ntips_basics = length(basics.tree$tip.label)

## get root node by finding node that isn't a child
root_fire = fire.tree$edge[(!fire.tree$edge[,1] %in% fire.tree$edge[,2]),1] %>% unique()
root_basics = basics.tree$edge[(!basics.tree$edge[,1] %in% basics.tree$edge[,2]),1] %>% unique()

## get max distance from root to another node (root age)
age_fire = max(dist.nodes(fire.tree)[,root_fire])
age_basics = max(dist.nodes(basics.tree)[,root_basics])

## rescale tree so that age of root is 1
fire.tree$edge.length = fire.tree$edge.length / age_fire
basics.tree$edge.length = basics.tree$edge.length / age_basics

## recalculate root_age to make sure things look good
age_fire = max(dist.nodes(fire.tree)[,root_fire])
age_basics = max(dist.nodes(basics.tree)[,root_basics])

## figure out nice printing things
label_offset_fire = 0.05 * age_fire
label_offset_basics = 0.05 * age_basics

tree_width_fire = 1.5 * age_fire
tree_width_basics = 1.5 * age_basics

# make sure both trees are set to rooted
attr(fire.tree, "rooted") <- TRUE
attr(basics.tree, "rooted") <- TRUE

# print tree and data if want
# attributes(fire.tree)
# is.rooted(fire.tree)
# fire.tree
# fire.data
# basics.tree
# basics.data
```


## Plotting the tree

```{r}

# create a bunch of pngs for basics trees which are otherwise too big to view (using plot.phylo())
png("basics_fan.png", width = 4000, height = 4000)
plot.phylo(basics.tree, type="fan", cex=1, edge.width = 5, 
           x.lim=c(-4, 10), y.lim=c(-4, 7))
dev.off()

png("basics_radial.png", width = 1000, height = 1000)
plot.phylo(basics.tree, type="radial", cex=1.5, edge.width = 5, )
dev.off()

png("basics_cladogram.png", width = 1000, height = 1000)
plot.phylo(basics.tree, type="cladogram", cex=1.5, edge.width = 5, )
dev.off()

png("basics_tidy.png", width = 1000, height = 1000)
plot.phylo(basics.tree, type="tidy", cex=1.5, edge.width = 5 )
dev.off()

    
png("basics_phylogram.png", width = 2000, height = 2000)
plot(basics.tree, cex=1.8)
nodelabels(basics.tree$node.label, cex=0.2)
dev.off()

# also show a small tidy basics tree in viewer
plot.phylo(basics.tree, type="tidy", cex=0.5, edge.width = 1 )

# ggtree plot for basics.tree
basics_ggtree = ggtree(basics.tree) + 
  geom_tiplab(fontface = "italic", size=2, offset = label_offset_basics/2) +
  xlim(0, tree_width_basics) +
  geom_text2(aes(label=node), col="red", size=2, nudge_x=0.005)
basics_ggtree

# ggtree plot for fire.tree
fire_ggtree = ggtree(fire.tree) + geom_tiplab(fontface = "italic",
  offset = label_offset_fire) +
  xlim(0, tree_width_fire) +
  geom_text2(aes(label=node), col="red", nudge_x=label_offset_fire/2)
fire_ggtree
```

## Ancestral State Reconstruction

```{r}
## ASR!
?ace

fire_hab_values <- fire.data$X.habitat.
basics_hab_values <- basics.data$X.habitat.
fire.tree$node.label
basics.tree$node.label

if (is.null(fire.tree$node.label) || any(fire.tree$node.label == "")) {
  fire.tree$node.label <- paste0("Node", 1:fire.tree$Nnode)
}
fire.tree$node.label <- NULL

if (is.null(basics.tree$node.label) || any(basics.tree$node.label == "")) {
  basics.tree$node.label <- paste0("Node", 1:basics.tree$Nnode)
}
basics.tree$node.label <- NULL

# print("fire.tree$node.label:")
# str(fire.tree$node.label)
# print("fire.tree$tip.label:")
# str(fire.tree$tip.label)
# print("fire.tree$edge:")
# str(fire.tree$edge)
# print("fire_hab_values:")
# str(fire_hab_values)

fire_habitat_ancestral = 
  ace (
    fire_hab_values,
    fire.tree,
    type="discrete", 
    method="ML",
    model="ER"
  )

basics_habitat_ancestral = 
  ace (
    basics_hab_values,
    basics.tree,
    type="discrete", 
    method="ML",
    model="ER"
  )

str(fire_habitat_ancestral)
names(fire_habitat_ancestral)

str(basics_habitat_ancestral)
names(basics_habitat_ancestral)

## why isn't it showing me these... :sob:
# '$ace' contains the reconstructed ancestral character states
# '$sigma2' is the estimate of the Brownian motion parameter
# '$CI95' are the confidence intervals on the ASR

```

### Map ancestral states onto phylogeny
```{r}

# get habitat data ready to map onto the phylogeny
ancestral_states_index <- apply(fire_habitat_ancestral$lik.anc, 1, which.max)
state_values <- c(2, 3, 4, 7)
ancestral_states <- state_values[ancestral_states_index]
fire_hab_values <- factor(fire_hab_values, levels = state_values) 
ancestral_states <- factor(ancestral_states, levels = state_values)

basics_ancestral_states_index <- apply(basics_habitat_ancestral$lik.anc, 1, which.max)
basics_state_values <- c(1, 2, 3, 4, 5, 6, 7, 8)
basics_ancestral_states <- state_values[basics_ancestral_states_index]
basics_hab_values <- factor(basics_hab_values, levels = basics_state_values) 
basics_ancestral_states <- factor(basics_ancestral_states, levels = basics_state_values)

# create and call plot with circles at each node
node_vals = c(fire_hab_values, ancestral_states)
basics_node_vals = c(basics_hab_values, basics_ancestral_states)

this_palette <- palette()
colors <- setNames(c(this_palette[1:length(state_values)]), state_values)
basics_colors <- setNames(c(this_palette[1:length(basics_state_values)]), basics_state_values)

legend_labels <- c("grassland", "mountain", "rare", "urban")
basics_legend_labels <- c("forest", "grassland", "mountain", "rare", "rough-terrain", "sea", "urban", "waters-edge")

fire_rect = ggtree(fire.tree) + geom_tiplab(fontface = "italic",
  offset = label_offset_fire) +
  xlim(0, tree_width_fire) +
  geom_tippoint(aes(color=node_vals), size=3, alpha=1) + 
  geom_nodepoint(aes(color=node_vals), size=3, alpha=1) + 
  scale_color_manual(
    values = colors, 
    labels = legend_labels,
    name = "habitat"
  )
fire_rect

basics_rect = ggtree(basics.tree) + geom_tiplab(fontface = "italic",
  offset = label_offset_basics) +
  xlim(0, tree_width_basics) +
  geom_tippoint(aes(color=basics_node_vals), size=3, alpha=1) + 
  geom_nodepoint(aes(color=basics_node_vals), size=3, alpha=1) + 
  scale_color_manual(
    values = basics_colors, 
    labels = basics_legend_labels,
    name = "habitat"
  )
basics_rect

?ggtree
# creating a plot where branches are colored by state
fire.tree$state <- as.factor(node_vals)
fire_branch_colors = ggtree(fire.tree) +
  geom_tree(aes(color = fire.tree$state), size = 2) +
  geom_tiplab(fontface = "italic", offset = 0.05) +
  xlim(0, tree_width_fire) +
  scale_color_manual(
    values = colors, 
    labels = legend_labels,
    name = "habitat"
  )
fire_branch_colors

basics.tree$state <- as.factor(basics_node_vals)
basics_branch_colors = ggtree(basics.tree) +
  geom_tree(aes(color = basics.tree$state), size = 2) +
  geom_tiplab(fontface = "italic", offset = 0.05) +
  xlim(0, tree_width_basics) +
  scale_color_manual(
    values = basics_colors, 
    labels = basics_legend_labels,
    name = "habitat"
  )
basics_branch_colors

# ^ but fan
fire_fan = ggtree(fire.tree, layout = "fan") +
  geom_tree(aes(color = fire.tree$state), size = 2) +
  geom_tiplab(fontface = "italic", offset = 0.05) +
  xlim(0, tree_width_fire) +
  scale_color_manual(
    values = colors, 
    labels = legend_labels,
    name = "habitat"
  )
fire_fan

basics_fan = ggtree(basics.tree, layout = "fan") +
  geom_tree(aes(color = basics.tree$state), size = 2) +
  geom_tiplab(fontface = "italic", offset = 0.05) +
  xlim(0, tree_width_basics) +
  scale_color_manual(
    values = basics_colors, 
    labels = basics_legend_labels,
    name = "habitat"
  )
basics_fan


## posterior probabilities as pie charts on internal nodes
?make.simmap

# create hab_states factor for tip states
hab_states <- state_values[as.integer(fire_hab_values)]
hab_states <- factor(hab_states, levels = state_values)
names(hab_states) <- fire.tree$tip.label

basics_hab_states <- basics_state_values[as.integer(basics_hab_values)]
basics_hab_states <- factor(basics_hab_states, levels = basics_state_values)
names(basics_hab_states) <- basics.tree$tip.label

print("smth")
str(basics_hab_states)

trees <- make.simmap(fire.tree, hab_states, model="ER", nsim=100)
# basics_trees <- make.simmap(basics.tree, basics_hab_states, model="ER", nsim=100)

obj <- summary(trees, plot=FALSE)
plot(obj, colors=colors, fsize=0.8, cex=c(0.5, 0.3))
?add.simmap.legend
add.simmap.legend(
  colors=colors, 
  x=0.9*par()$usr[1],
  y=0.3*par()$usr[4],
  prompt=FALSE,
  fsize=0.9, 
  leg = legend_labels
)

# basics_obj <- summary(basics_trees, plot=FALSE)
# plot(basics_obj, colors=basics_colors, fsize=0.8, cex=c(0.5, 0.3))
# add.simmap.legend(
#   colors=basics_colors, 
#   x=0.9*par()$usr[1],
#   y=0.3*par()$usr[4],
#   prompt=FALSE,
#   fsize=0.9, 
#   leg = basics_legend_labels
# )

# ARD model
legend_labels <- c("2: grassland", "3: mountain", "4: rare", "7: urban")
basics_legend_labels <- c("1: forest", "2: grassland", "3: mountain", "4: rare", 
                          "5: rough-terrain", "6: sea", "7: urban", "8: waters-edge")


fit.ARD<-fitMk(fire.tree, hab_states, model="ARD")
plot(fit.ARD, show.zeros=FALSE, main="fitted ARD model for habitat evolution")
legend("bottomright", legend = legend_labels, title = "habitats")

# basics_fit.ARD<-fitMk(basics.tree, basics_hab_states, model="ARD")
# plot(basics_fit.ARD, show.zeros=FALSE, main="fitted ARD model for basics habitat evolution")
# legend("bottomright", legend = basics_legend_labels, title = "basics habitats")
```

# Things I Would Love To Do But I Tried For a While And Got Nowhere
```{r}

# # cool but only when already know asr states :eyeroll:
# # ## for discrete states (ANOLE EXAMPLE)
# data(anoletree)
# cols<-setNames(palette()[1:6],mapped.states(anoletree))
# # cols <- setNames(palette()[1:length(unique(X))], sort(unique(X)))
# 
# plot(anoletree, cols, type="fan", fsize=0.8, lwd=3, ftype="i")
# add.simmap.legend(colors=cols, x = 0.9*par()$usr[1],
#     y = 0.9*par()$usr[4], prompt=FALSE, fsize=0.9)
# 
# ## figure out what mapped.states() is doing
# ?mapped.states
# str(anoletree)
# result <- mapped.states(anoletree)
# print("result:")
# print(result)
# 
# # plot discrete tip node states as colors
# plotTree(fire.tree, type="fan", fsize=0.8, ftype="i")
# cols <- setNames(palette()[1:length(unique(X))], sort(unique(X)))
# tiplabels(pie=to.matrix(X, sort(unique(X))), piecol = cols, cex = 0.3)
# add.simmap.legend(colors=cols,prompt=FALSE,x=0.9*par()$usr[1],
#     y=-max(nodeHeights(tree)),fsize=0.5)
```


# Interesting Results
Legendary pokémon Moltres *failed* the chi2 test when compared with all the other fire gen1 pokémon!
                      Gap/Ambiguity  Composition  p-value
   1  charmander              0.00%    passed     61.69%
   2  charmeleon_charizard    0.00%    passed     69.95%
   3  vulpix                  0.00%    passed     88.29%
   4  ninetales               0.00%    passed     98.01%
   5  growlithe_arcanine      0.00%    passed     88.29%
   6  ponyta_rapidash         0.00%    passed     89.76%
   7  magmar                  0.00%    passed     24.30%
   8  flareon                 0.00%    passed     77.79%
   9  moltres                 0.00%    failed      3.34%
****  TOTAL                   0.00%  1 sequences failed composition chi2 test (p-value<5%; df=11)


