---
title: "Ancestral State Reconstruction"
output:
  pdf_document: default
  html_document: default
---

```{r setup, echo = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# for large palettes
library(pals)

# General data analysis packages
library(knitr)
library(tidyverse)
library(magrittr)
library(plotrix)
library(plotly)

# Phylogenetics packages
library(ape)
library(phylobase)
library(phangorn)
library(phytools)
library(tidyverse)
library(ggtree)
library(ggplot2)

```

based on (this tutorial)[http://www.phytools.org/eqg2015/asr.html].

# Set file to analyze

```{r}
# read in tree and csv, set filename for folder/file creation
name = "basic"

# choose whether to root tree at midpoint or choose a specific tip node as root (ie "mew")
root_choice = "mew" # set it to "midpoint" for rooting at midpoint, and "mew" for rooting at mew (and anything else for rooting at anything else!)

# define features to reconstruct ancestral states for
features <- c("type", "ability", "stat", "shape", "color", "habitat", "egg_group", "region")
```


# Clean files

```{r}
# read in files
tree<-read.tree(sprintf("%s/%s.treefile", name, name))
data<-read.csv(sprintf("%s/%s.csv", name, name), row.names = 1)

# bookkeeping
## (checked for duplicate tips in python before creating .nex and .csv files!)

## make sure rows of character data are in same order as tip labels
data = data[order(match(row.names(data), tree$tip.label)), ]

## double check that labels are the same, in case some were missing or duplicated
if(! all( row.names(data)==tree$tip.label ) ){
  stop("fire.tree tip labels and dataframe rows not in same order!")
}

## resolve polytomies
tree <- multi2di(tree)

## set root tree based on previous rooting choice
if (root_choice == "midpoint"){
  tree <- midpoint(tree)
} else {
  tree <- root(tree, root_choice, resolve.root = TRUE)
}

# make sure tree is set to rooted so that later analyses will work
attr(tree, "rooted") <- TRUE

# find and set important metadata abt each tree
## count number of tips
n_tips = length(tree$tip.label)

## get root node by finding node that isn't a child
root_node = tree$edge[(!tree$edge[,1] %in% tree$edge[,2]),1] %>% unique()

## get max distance from root to another node (root age)
root_age = max(dist.nodes(tree)[,root_node])

## rescale tree so that age of root is 1
tree$edge.length = tree$edge.length / root_age

## recalculate root_age to make sure things look good
root_age = max(dist.nodes(tree)[,root_node])

## find good label offset and tree width
label_offset = 0.05 * root_age
tree_width = 1.5 * root_age

# print tree and data if want
attributes(tree)
is.rooted(tree)
str(tree)
data
print("data:")
str(data)
```


## Plotting the tree

```{r}

# create visualizations and save to visualizations folder
png(sprintf("%s/visualizations/fan.png", name), width = 4000, height = 4000)
plot.phylo(tree, type="fan", cex=1, edge.width = 5)
dev.off()

png(sprintf("%s/visualizations/radial.png", name), width = 4000, height = 4000)
plot.phylo(tree, type="radial", cex=1, edge.width = 5, )
dev.off()

png(sprintf("%s/visualizations/cladogram.png", name), width = 1000, height = 1000)
plot.phylo(tree, type="cladogram", cex=1, edge.width = 5, )
dev.off()

png(sprintf("%s/visualizations/tidy.png", name), width = 2000, height = 2000)
plot.phylo(tree, type="tidy", cex=0.5, edge.width = 5 )
dev.off()

png(sprintf("%s/visualizations/phylogram.png", name), width = 2000, height = 2000)
plot(tree, cex=1)
nodelabels(tree$node.label, cex=0.2)
dev.off()

# also show a small tidy tree in viewer
plot.phylo(tree, type="tidy", cex=0.5, edge.width = 1 )

# ggtree plot
png(sprintf("%s/topology.png", name), width = 2000, height = 2000)
gg_tree = ggtree(tree) + 
  geom_tiplab(fontface = "italic", offset = 0.0005) +
  xlim(0, tree_width)
gg_tree
dev.off()
```

## Ancestral State Reconstruction

```{r}
# initialize list to store results in
ancestral_results <- list()
tree$node.label <- NULL # set all internal nodes to null initially (so that ace can set them?)

# for each feature to reconstruct:
for (feature in features){
  
  # reconstruct feature
  str(feature)
  print("tree:")
  str(tree)
  
  column_name <- paste0("X.", feature, ".") # add X. prefix
  curr_data <- data[[column_name]] # get tip data for feature
  ancestral_results[[feature]] <- # perform ACE
    ace(curr_data, tree, type="discrete", method="ML", model="ER")
  
  print("ancestral_results:")
  str(ancestral_results[[feature]])
  
  print("...$lik.anc:")
  str(ancestral_results[[feature]]$lik.anc)
  
  # map data onto phylogeny
  
  ## adjust for potential missing possible states by adding empty columns to likelihood matrix
  correct_cols <- c('0', '1', '2', '3', '4', '5', '6', '7', '8', '9', 
                  'A', 'B', 'C', 'D', 'E', 'F', 'G', 'H', 'I', 'J', 
                  'K', 'L', 'M', 'N', 'O', 'P', 'Q', 'R', 'S', 'T', 
                  'U', 'V', 'W', 'X', 'Y', 'Z')
  current_cols <- colnames(ancestral_results[[feature]]$lik.anc) # extract current column names
  last_curr_col <- tail(current_cols, 1) # find final column
  valid_cols <- correct_cols[1:which(correct_cols == last_curr_col)] # create subset of correct_cols, going up to found final column
  missing_cols <- setdiff(valid_cols, current_cols) # find missing cols
  if (length(missing_cols) > 0){ # if missing cols exist
      for (col in missing_cols){ 
          new_col <- matrix(-Inf, nrow = nrow(ancestral_results[[feature]]$lik.anc), ncol = 1, 
                            dimnames = list(NULL, col))
            # add col at end with name of missing col
            # initializing all values to negative infinity as to not get chosen
          ancestral_results[[feature]]$lik.anc <- cbind(ancestral_results[[feature]]$lik.anc, new_col) 
      }                                                        
      # re-order columns to align with valid_cols order
      ancestral_results[[feature]]$lik.anc <- ancestral_results[[feature]]$lik.anc[, valid_cols]
  }
  
  ## get indices (without ^, would be potentially incorrect indices)
  ancestral_state_indices <- apply(ancestral_results[[feature]]$lik.anc, 1, which.max) 
    # returns vector of indices length of pokémons, where each index is the index of the highest-likelihood state for this feature
  
  ## get tip values, also adjusting for missing values
  tip_states <- unique(data[[column_name]]) # gets all seen tip states for this feature
  tip_states <- tip_states[order( # order list 0-9 and then A-Z
      as.numeric(grepl("[^0-9]", tip_states)), # numbers first
      tip_states # then alphabetically
  )]
  last_tip_state <- tail(tip_states, 1) # find final tip state seen
  state_values <- correct_cols[1:which(correct_cols == last_tip_state)] # create subset of correct_cols, going up to found final states
  
  print("ancestral state indices:")
  str(ancestral_state_indices)
  
  print("state values:")
  str(state_values)
  
  
  curr_data <- factor(curr_data, levels = state_values) # ensures order of levels is same btwn data and predictions
  ancestral_states <- state_values[ancestral_state_indices] # returns vector of highest-likelihood state for each pokemon
  ancestral_states <- factor(ancestral_states, levels = state_values) # factors this vector by state values
  
  print("curr_data:")
  str(curr_data)
  
  # CURR DATA IS RIGHT
  
  print("ancestral_states:")
  str(ancestral_states)
  
  # SO ANCESTRAL STATES ARE PROBABLY RIGHT TOO AND IT'S THE SAME PROBLEM JUST LATER
  

  # set up plot
  ## combine known tip states with predicted internal node states
  node_values = c(curr_data, ancestral_states) 
  print("node_values:")
  str(node_values)
  
  print("state_values:")
  str(state_values)
  
  ## colors
  c25 <- c(
    "dodgerblue2", "#E31A1C", # red
    "green4",
    "#6A3D9A", # purple
    "#FF7F00", # orange
    "black", "gold1",
    "skyblue2", "#FB9A99", # lt pink
    "palegreen2",
    "#CAB2D6", # lt purple
    "#FDBF6F", # lt orange 
    "gray70", "khaki2",
    "maroon", "orchid1", "deeppink1", "blue1", "steelblue4",
    "darkturquoise", "green1", "yellow4", "yellow3",
    "darkorange4", "brown"
  )
  this_palette <- c25
  colors <- setNames(c(this_palette[1:length(state_values)]), state_values) # assigns each state value a colors
  ### if feature is pokemon color, change colors to match
  color_vector = c("black", "blue", "brown", "gray", "green", "pink", "purple", "red", "white", "yellow")
  type_vector <- c("pink", "#C22E28", "blue1", "#A33EA1", "darkorange4", "gray35", "#A6B91A", "#BFA1CD", "#B7B7CE", "#EE8130",
          "#6390F0", "#7AC74C", "#F7D02C", "#6F35FC", "#96D9D6", "darkturquoise", "black", "deeppink1")
  if (feature == "color"){
      colors <- setNames(color_vector, state_values)
  } else if (feature == "type"){
      colors <- setNames(type_vector, state_values)
  }
  print("colors:")
  str(colors)
  print(colors)
  
  ## labels depend on feature
  legend_labels <- switch(
    feature,
    "type"= c("normal", "fighting", "flying", "poison", "ground", "rock", "bug", "ghost", "steel", 
              "fire", "water", "grass", "electric", "psychic", "ice", "dragon", "dark", "fairy"), 
    "ability"= 1:18, 
    "stat"= c("hp", "attack", "defense", "special-attack", "special-defense", "speed"),
    "shape"= c("ball", "squiggle", "fish", "arms", "blob", "upright", "legs", "quadruped", "wings", "tentacles", "heads",
               "humanoid", "bug-wings", "armor"),
    "color"= color_vector,
    "habitat"= c("forest", "grassland", "mountain", "rare", "rough-terrain", "sea", "urban", "waters-edge", "unknown"),
    "egg_group"= c("monster", "water1", "bug", "flying", "ground", "fairy", "plant", "humanshape", "water3",
                   "mineral", "indeterminate", "water2", "ditto", "dragon", "no-eggs"),
    "region"= c("Unknown", "Kanto", "Johto", "Hoenn", "Sinnoh", "Unova")
  )
  
  print("legend labels:")
  str(legend_labels)
  print(legend_labels)
  
  # MAKE VISUALIZATIONS
  ## rectangular plot
  png(sprintf("%s/ASR/%s/rect.png", name, feature), width = 2000, height = 2000)
  rect = ggtree(tree) +
      geom_tiplab(fontface = "italic",
      offset = 0.0005) +
      xlim(0, tree_width) +
      geom_tippoint(aes(color=node_values), size=3, alpha=1) +
      geom_nodepoint(aes(color=node_values), size=3, alpha=1) +
      scale_color_manual(
          values = colors,
          breaks = names(colors), 
          labels = legend_labels,
          name = feature
      )
  print(rect)
  dev.off()
  
  ## rectangular plot where branches are colored by state
  png(sprintf("%s/ASR/%s/rect_branch_colors.png", name, feature), width = 2000, height = 2000)
  tree$state <- as.factor(node_values)
  rect_branch_colors <- ggtree(tree) +
      geom_tree(aes(color = tree$state), size=1) +
      geom_tiplab(fontface = "italic", offset = 0.0005) +
      xlim(0, tree_width) +
      scale_color_manual(
        values = colors, 
        breaks = names(colors), 
        labels = legend_labels,
        name = feature
      )
  print(rect_branch_colors)
  dev.off()
  
  ## fan branch color plot
  png(sprintf("%s/ASR/%s/fan_branch_colors.png", name, feature), width = 2000, height = 2000)
  fan_branch_colors <- ggtree(tree, layout="fan") +
      geom_tree(aes(color = tree$state), size=1) +
      geom_tiplab(fontface = "italic", offset = 0.0005) +
      xlim(0, tree_width) +
      scale_color_manual(
        values = colors, 
        breaks = names(colors), 
        labels = legend_labels,
        name = feature
      )
  print(fan_branch_colors)
  dev.off()
  
  # # posterior probabilities as pie charts on internal nodes
  # ## create states factor for tip states
  # png(sprintf("%s/ASR/%s/simmap_plot.png", name, feature), width = 2000, height = 2000)
  # states <- factor(state_values[as.integer(curr_data)], levels = state_values)
  # names(states) <- tree$tip.label
  # 
  # trees <- make.simmap(tree, states, model="ER", nsim=100)
  # obj <- summary(trees, plot=FALSE)
  # plot(obj, colors=colors, fsize=0.8, cex=c(0.1, 0.3))
  # add.simmap.legend(
  #     colors=colors,
  #     x=0.9*par()$usr[1],
  #     y=0.3*par()$usr[4],
  #     prompt=FALSE,
  #     fsize=0.9,
  #     leg = legend_labels
  # )
  # dev.off()
  #
  # # ARD model
  # png(sprintf("%s/ASR/%s/ARD.png", name, feature), width = 2000, height = 2000)
  # fit.ARD <- fitMk(tree, states, model="ARD")
  # plot(fit.ARD, show.zeros=FALSE, main="fitted ARD model")
  # legend("bottomright", legend = legend.labels, title = feature)
  # dev.off()

}

## why isn't it showing me these... :sob:
# '$ace' contains the reconstructed ancestral character states
# '$sigma2' is the estimate of the Brownian motion parameter
# '$CI95' are the confidence intervals on the ASR

```

# Things I Would Love To Do But I Tried For a While And Got Nowhere
```{r}

# # cool but only when already know asr states :eyeroll:
# # ## for discrete states (ANOLE EXAMPLE)
# data(anoletree)
# cols<-setNames(palette()[1:6],mapped.states(anoletree))
# # cols <- setNames(palette()[1:length(unique(X))], sort(unique(X)))
# 
# plot(anoletree, cols, type="fan", fsize=0.8, lwd=3, ftype="i")
# add.simmap.legend(colors=cols, x = 0.9*par()$usr[1],
#     y = 0.9*par()$usr[4], prompt=FALSE, fsize=0.9)
# 
# ## figure out what mapped.states() is doing
# ?mapped.states
# str(anoletree)
# result <- mapped.states(anoletree)
# print("result:")
# print(result)
# 
# # plot discrete tip node states as colors
# plotTree(fire.tree, type="fan", fsize=0.8, ftype="i")
# cols <- setNames(palette()[1:length(unique(X))], sort(unique(X)))
# tiplabels(pie=to.matrix(X, sort(unique(X))), piecol = cols, cex = 0.3)
# add.simmap.legend(colors=cols,prompt=FALSE,x=0.9*par()$usr[1],
#     y=-max(nodeHeights(tree)),fsize=0.5)





# flying types don't exist??????
```


# Interesting Results
Legendary pokémon Moltres *failed* the chi2 test when compared with all the other fire gen1 pokémon!
                      Gap/Ambiguity  Composition  p-value
   1  charmander              0.00%    passed     61.69%
   2  charmeleon_charizard    0.00%    passed     69.95%
   3  vulpix                  0.00%    passed     88.29%
   4  ninetales               0.00%    passed     98.01%
   5  growlithe_arcanine      0.00%    passed     88.29%
   6  ponyta_rapidash         0.00%    passed     89.76%
   7  magmar                  0.00%    passed     24.30%
   8  flareon                 0.00%    passed     77.79%
   9  moltres                 0.00%    failed      3.34%
****  TOTAL                   0.00%  1 sequences failed composition chi2 test (p-value<5%; df=11)


