---
title: "exercise 10"
output: html_document
---

```{r setup, echo = FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# General data analysis packages
library(knitr)
library(tidyverse)
library(magrittr)

# Phylogenetics packages
library(ape)
library(phylobase)
library(phytools)
library(tidyverse)
library(ggtree)

```

# From the Exercise:

- Important R packages for phylogenetic analysis: on [this list](https://cran.r-project.org/web/views/Phylogenetics.html). 
- Use "Run->Rin all" from the menu for debugging, then type commands into the console when it stops
- knit/run often. identify incremental goals and test progress each time

# Important Lines of Code:
```{r}
# make sure rows of character data are in same order as tip labels
geoData = geoData[ match( row.names(geoData), geoTree$tip.label  ), ]

# Double check that labels are the same, in case some were missing or duplicated
if(! all( row.names(geoData)==geoTree$tip.label ) ){
  stop("Error: tip labels and dataframe rows not in same order!")
}

## make look good / ultrametric?:

# Count the number of tips
n_tips = length(geoTree$tip.label)

# Get the root node by finding the node that is a parent (in column 1 of edge) 
# but not a child (in column 2 of edge)
root_node = geoTree$edge[(!geoTree$edge[,1] %in% geoTree$edge[,2]),1] %>% unique()

# Get the maximum distance from the root to another node, which will be for a tip.
# Since this tree happens to be ultrametric, this will be the root age
root_age = max( dist.nodes(geoTree)[,root_node] )

# This isn't necessary, but since we don't know the units of edge length let's rescale
# the tree so that the age of the root is 1.

geoTree$edge.length = geoTree$edge.length / root_age

# And now recalculate the root age to make sure things look good
root_age = max( dist.nodes(geoTree)[,root_node] )

# kable prints pretty tables
kable(geoData)


## show phylogeny w node numbers

# First, get some values that will help us scale the tree on the page well. Would 
# by nice if ggtree did that, but it doesn't.

label_offset = 0.05 * root_age
tree_width = 1.5 * root_age

geo_ggtree = ggtree( geoTree ) + 
  geom_tiplab( fontface = "italic", offset=label_offset ) +
  xlim(0, tree_width) +
  geom_text2(aes(label=node), col="red", nudge_x=label_offset/2 )

geo_ggtree

```

```{r}

## plot node trait values
# To display the node values, ggtree expects a vector where each element
# corresponds to each node in the tree, both at the tips and at the internal 
# nodes. Since we don't have values for the internal nodes, we will pad them 
# with NA (Not Available) for now.
node_values = c( geoData$wingL, rep(NA, n_tips-1) )

geo_ggtree + 
  geom_tippoint( aes(color=node_values), size=3, alpha=1 )

```

## Ancestral character state estimation

Given the phylogeny (topology and edge lengths) and character data, we can now reconstruct some ancestral states. We will use the ape function `ace()`, consult the [ape manual](https://cran.r-project.org/web/packages/ape/ape.pdf) for more information.


```{r}

str(geoData$wingL)
str(geoTree)

wing_ancestral = 
  ace( 
    geoData$wingL, 
    geoTree, 
    type="continuous", 
    method="REML", 
    model="BM"  
  )

wing_ancestral

```

This gives us a variety of information about the ancestral character reconstructions:

- `$ace` contains the reconstructed ancestral character states. There is one value per internal node, and they are in the order that the internal nodes are numbered in.

- `$sigma2` is the estimate of the Brownian Motion parameter.

- `$CI95` are the confidence intervals on the ancestral character state reconstructions.


Let's plot the reconstructed states on the phylogeny:


```{r}

# To display the node values, ggtree expects a vector where each element
# corresponds to each node in the tree, both at the tips and at the internal 
# nodes. Since we don't have values for the internal nodes, we will pad them 
# with NA (Not Available) for now.
node_values = c( geoData$wingL, wing_ancestral$ace )

geo_ggtree + 
  geom_tippoint( aes(color=node_values), size=3, alpha=1 ) +
  geom_nodepoint(  aes(color=node_values), size=3, alpha=1 )

```

Above we used the REML (REduced Maximum-Likelihood) method of inference, an approach developed by Joe Felsenstein that mitigates problems with edge lengths under ML for continuous traits.

An alternative approach is the averaging method used in Phylogenetic Independent Contrasts, also developed by Joe Felsenstein.

```{r}

str(geoData$wingL)
str(geoTree)

wing_ancestral_pic = 
  ace( 
    geoData$wingL, 
    geoTree, 
    type="continuous", 
    method="pic", 
    model="BM"  
  )

wing_ancestral_pic

```


Let's compare the methods more directly. Here are the ancestral character state reconstructions, as a scatterplot. I've included the numbers of the nodes, as well as a blue line that indicates identical values.


```{r}

D_methods = data.frame(
  reml = wing_ancestral$ace,
  pic = wing_ancestral_pic$ace
)

# Add node numbers to the data frame
D_methods$node = seq(1:nrow(D_methods)) + n_tips

D_methods %>%
  ggplot(aes(x=reml,y=pic)) +
  geom_point() +
  geom_text( aes(label=node), color="red", vjust = 0, nudge_y = 0.005 ) +
  geom_abline(slope=1, intercept=0, color="blue")

```

## Confidence and node age

The code below explores the width of the confidence intervals for each reconstruction.

```{r}

interval_width = wing_ancestral$CI95[,2] - wing_ancestral$CI95[,1]

node_values = c( rep(NA, nrow(geoData)), interval_width )

kable( data.frame( node = seq(1, length(node_values)), width=node_values ) )


geo_ggtree + 
  geom_nodepoint(  aes(color=node_values), size=3, alpha=1 )


```


## Simulation

The code below simulates a new dataset under Brownian Motion on the same tree.

```{r}

  set.seed(42) # Use the same seed value for the simulation each time it is run
  sim_x = fastBM(geoTree, a=2, sig2=0.5, internal = TRUE)
  D_sim = data.frame( node = seq(1, length(sim_x)), sim_x=sim_x )
  kable( D_sim )

  geo_ggtree + 
    geom_tippoint( aes(color=sim_x), size=3, alpha=1 ) +
    geom_nodepoint(  aes(color=sim_x), size=3, alpha=1 )
```


Now let's try some ancestral character state reconstructions, and compare them to the known simulated values.

```{r}

tip_x = sim_x[1:n_tips]

reconstruction_ancestral_x = 
  ace( 
    tip_x, 
    geoTree, 
    type="continuous", 
    method="REML", 
    model="BM"  
  )

reconstruction_ancestral_x

D_sim$reconstruction = NA

D_sim$reconstruction[(n_tips+1):nrow(D_sim)] = reconstruction_ancestral_x$ace


D_sim[(n_tips+1):nrow(D_sim),] %>%
  ggplot(aes(x=sim_x,y=reconstruction)) +
  geom_point() +
  geom_text( aes(label=node), color="red", vjust = 0, nudge_y = 0.005 ) +
  geom_abline(slope=1, intercept=0, color="blue")

```

# My own

from [here](http://www.phytools.org/eqg2015/asr.html).

```{r}

```







